Meta-analytic approaches and effect sizes to account for 'nuisance heterogeneity' in comparative physiology

### 2021-09-18 MS/ms.Rmd
Fixed more minor revisions. Need to add a couple more refs

Meta-analysis is a powerful tool to generate quantitatively informed answers to pressing global challenges. By distilling data from broad sets of research designs and study systems into standardized effect sizes, meta-analyses provide physiologists with opportunities to estimate overall effect sizes and understand the drivers of effect variability. Despite this ambition, research designs in the field of comparative physiology can appear, at the outset, as being vastly different to each other because of 'nuisance heterogeneity' (e.g., different temperatures or treatment dosages used across studies). Methodological differences across studies have led many to believe that meta-analysis is an exercise in comparing "apples with oranges". Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with multilevel meta-regression models to both account for the factors driving differences across studies and make them more comparable. We formalise effect size measures (e.g., $Q_{10}$) that provide comparative physiologists with a means to remove nuisance heterogeneity without the need to resort to more complex statistical models that may be harder to interpret. We also describe more general approaches that can be applied to a variety of different contexts to derive new effects sizes and sampling variances opening new possibilities. By using effect sizes that account for components of effect heterogeneity, in combination with existing meta-analytic models, comparative physiologists can explore exciting new questions and make results from large-scale data sets more accessible, comparable and widely interpretable.

@@

As can be seen in Table \@ref(tab:tablea), none of the effect measures control for nuisance heterogeneity. For example, the magnitude of *lnRR* or *SMD* might depend on differences in the temperatures, dosages, pH, or water potential applied to each treatment. The same applies to modelling the means. For example, standard metabolic rate (SMR) is temperature-dependent. As such, one needs to standardise the temperature measurement to compare across studies [@White2006; @Uyeda2017] (if, of course, temperature is not of interest itself). While we could model this explicitly, this may not always be desirable. It can often be useful to standardise the effect to both ease its interpretation and simplify modelling. In Box 1, we provide readers with guidance on how comparative physiologists can apply corrections to the effect measure itself. We highlight two examples in Box 1 that can be easily generalized to alternative effect size measures and cases. Below, we show how we can formalise a traditional effect size, such as $lnRR$, to capture this nuisance variation using a common, well understood effect measure of interest to comparative physiologists ($Q_{10}$). Figure \@ref(fig:fig0) summarises the different effect sizes that we mention in Table 1, Box 1 and the section below.

>>>>

Meta-analysis is a powerful tool to generate quantitatively informed answers to pressing global challenges. By distilling data from broad sets of research designs and study systems into standardized effect sizes, meta-analyses provide physiologists with opportunities to estimate overall effect sizes and understand the drivers of effect variability. Despite this ambition, research designs in the field of comparative physiology can appear, at the outset, as being vastly different to each other because of 'nuisance heterogeneity' (e.g., different temperatures or treatment dosages used across studies). Methodological differences across studies have led many to believe that meta-analysis is an exercise in comparing "apples with oranges". Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with multilevel meta-regression models to both account for the factors driving differences across studies and make them more comparable. We formalise effect size measures (e.g., the temperature coefficient, $Q_{10}$) that provide comparative physiologists with a means to remove nuisance heterogeneity without the need to resort to more complex statistical models that may be harder to interpret. We also describe more general approaches that can be applied to a variety of different contexts to derive new effects sizes and sampling variances opening new possibilities. By using effect sizes that account for components of effect heterogeneity, in combination with existing meta-analytic models, comparative physiologists can explore exciting new questions and make results from large-scale data sets more accessible, comparable and widely interpretable.

@@

As can be seen in Table \@ref(tab:tablea), none of the effect measures control for nuisance heterogeneity. For example, the magnitude of *lnRR* or *SMD* might depend on differences in the temperatures, dosages, or pH applied to each treatment. The same applies to modelling the means. For example, standard metabolic rate (SMR) is temperature-dependent. As such, one needs to standardise the temperature measurement to compare across studies [@White2006; @Uyeda2017] (if, of course, temperature is not of interest itself). While we could model this explicitly, this may not always be desirable. It can often be useful to standardise the effect to both ease its interpretation and simplify modelling. In Box 1, we provide readers with guidance on how comparative physiologists can apply corrections to the effect measure itself. We highlight two examples in Box 1 that can be easily generalized to alternative effect size measures and cases. Below, we show how we can formalise a traditional effect size, such as $lnRR$, to capture this nuisance variation. We use a common, well understood, effect measure of interest to comparative physiologists -- the temperature coefficient ($Q_{10}$). Figure \@ref(fig:fig0) summarises the different effect sizes that we mention in Table 1, Box 1 and the section below.
